---
title: "Introduction to polean"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to polean}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(polean)
```

For a brief package-level overview, see `?"polean-package"`. For the
class and constructor, see `?polean`.

## Motivation

R's logical type has three values: `TRUE`, `FALSE`, and `NA`. The
third is propagating: any expression involving `NA` is itself `NA`,
which is usually the right thing for missing data but is the wrong
thing when `NA` is being used as a placeholder for *something that
resembles a logical but should not propagate*.

A few examples:

- A flag derived from a numeric threshold where the boundary case is
  ambiguous, but you still want downstream logical operations to work.
- A boolean reduction of a column where you want to preserve the
  distinction between "genuinely true" and "true but with caveats."
- Any case where you would otherwise reach for a 4-state truth value
  (true, false, neither, both) but want the extra states to behave
  numerically like ordinary booleans.

`polean` introduces a non-propagating extension of the logical type
that preserves numerical equivalence with base R while adding two new
literals.

## The five literals

Every polean is one of five values:

| Literal      | Orientation | Strictness  |
|--------------|-------------|-------------|
| `POL_TRUE`   | true        | strict      |
| `POL_FALSE`  | false       | strict      |
| `POL_1`      | true        | not strict  |
| `POL_0`      | false       | not strict  |
| `POL_NA`     | NA          | —           |

`POL_TRUE` and `POL_1` share the same orientation (true) but differ in
strictness. Likewise for `POL_FALSE` and `POL_0`. Internally, every
polean is stored as a single complex number: the real part is the
orientation (0, 1, or NA) and the imaginary part is the strictness
flag (0 for strict, 1 for not strict).

The five values are exported as locked length-1 package objects of
those names and can be used directly:

```{r}
POL_1
POL_TRUE
POL_NA
```

## Two constructors

There are two constructors. They differ in whether the strictness
distinction is decided by an explicit flag (`polean()`) or by the R
type of the input (`as.polean()`).

### `polean(orientation, strict = FALSE, dim = NULL)`

The basic constructor. It does *not* dispatch on the R type of
`orientation`: strictness is set entirely by `strict`. The default
`strict = FALSE` produces not-strict values; pass `TRUE` for strict.

```{r}
polean(TRUE)                         # POL_1   — default strict = FALSE
polean(1)                            # POL_1   — same; type is ignored
polean(c(1, 0, NA))                  # POL_1 POL_0 POL_NA
polean(TRUE, strict = TRUE)          # POL_TRUE
polean(1, strict = TRUE)             # POL_TRUE
polean(1, dim = c(2, 3))             # 2x3 array of POL_1
```

### `as.polean(x)`

Type-dispatched coercion. The R type of `x` determines whether the
result is strict or not strict:

| Input type                        | Behavior                                |
|-----------------------------------|-----------------------------------------|
| `logical`                         | strict                                  |
| `integer`, `double`, `numeric`    | not strict                              |
| `complex`                         | strict if `Im(x) == 0`, else not strict |
| `character`                       | parse element-wise, then re-dispatch    |
| `list`                            | element-wise coercion                   |

```{r}
as.polean(TRUE)                  # POL_TRUE
as.polean(1)                     # POL_1
as.polean(1+0i)                  # POL_TRUE
as.polean(1+1i)                  # POL_1
as.polean("TRUE")                # POL_TRUE
as.polean("1+0i")                # POL_TRUE
as.polean("1.5")                 # POL_1
as.polean(list(TRUE, 1, NA))     # POL_TRUE POL_1 POL_NA
```

### Why does `as.polean(1)` produce `POL_1`, not `POL_TRUE`?

A numeric `1` in user code is rarely meant as a boolean literal, but
it is often *used* in boolean-adjacent contexts (e.g. `sum(x == y)`,
flags from a 0/1 indicator vector). Treating it as not strict marks
the fact that the user passed a number, not a logical, while keeping
the numerical comparison `as.polean(1) == 1` true. To get a strict
`POL_TRUE` from numeric input, either coerce to logical first
(`as.polean(as.logical(1))`), pass a real-valued complex
(`as.polean(1+0i)`), or use the constructor with the explicit flag
(`polean(1, strict = TRUE)`).

## Accessors

```{r}
x <- c(POL_TRUE, POL_1, POL_FALSE, POL_NA)
value(x)         # underlying complex encoding
name(x)          # printed labels
orientation(x)   # truth orientation as logical
strict(x)        # strict (TRUE) vs not strict (FALSE)
```

The return value of `strict()` on a `POL_NA` element is
implementation-defined; treat it as undefined.

## Two equality regimes

```{r}
POL_TRUE == POL_TRUE                 # TRUE  — same orientation and strictness
POL_TRUE == POL_1                    # FALSE — differ in strictness
POL_1    == 1                        # TRUE  — numerical equivalence
POL_TRUE == 1                        # TRUE  — numerical equivalence
POL_TRUE == TRUE                     # TRUE  — full encoding matches (1+0i == 1+0i)
POL_1    == TRUE                     # FALSE — full encoding differs (1+1i != 1+0i)
```

Equality has two regimes:

- **Encoded** (`polean == polean`, `polean == logical`, `polean == complex`):
  compares the full complex encoding, so both orientation and
  strictness must agree.
- **Numerical** (`polean == numeric`): compares only orientation
  (`Re(value) == e2`), so a not-strict and a strict polean of the same
  orientation both equal the same number.

The asymmetry preserves the invariant `as.polean(x) == x` for numeric
`x`, while still letting users distinguish strict from not strict
whenever a polean is on both sides — or the other side is logical or
complex.

## Combination

`c()` preserves orientation and strictness for polean inputs, and
applies `as.polean()` element-wise to non-polean inputs.

```{r}
c(POL_TRUE, POL_1, NA)               # POL_TRUE POL_1   POL_NA
c(POL_TRUE, 1, FALSE)                # POL_TRUE POL_1   POL_FALSE
c(polean(TRUE), 1, FALSE)            # POL_1    POL_1   POL_FALSE
```

The contrast between the second and third examples is just the
constructor difference: `polean(TRUE)` is not strict (constructor
default), while `POL_TRUE` is strict.

## Display

```{r}
polean(c(TRUE, FALSE, NA))             # POL_1 POL_0 POL_NA          — constructor default: not strict
as.polean(c(TRUE, FALSE, NA))          # POL_TRUE POL_FALSE POL_NA   — logical input -> strict
as.polean(c(1, 0, NA))                 # POL_1 POL_0 POL_NA          — numeric input -> not strict
c(POL_TRUE, POL_1, NA, POL_FALSE)
```

## Status

This package is in early development. The current release covers
construction, coercion, equality, and combination. Logical operators
(`&`, `|`, `!`) and arithmetic interactions are not yet defined; their
semantics under the strictness split are still being worked out.
