---
title: "Introduction to polean"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to polean}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(polean)
```

For a brief package-level overview, see `?"polean-package"`. For the
class and constructor, see `?polean`.

## Motivation

R's logical type has three values: `TRUE`, `FALSE`, and `NA`. The third
is propagating: any expression involving `NA` is itself `NA`, which is
usually the right thing for missing data but is the wrong thing when
`NA` is being used as a placeholder for *something that resembles a
logical but should not propagate*.

A few examples:

- A flag derived from a numeric threshold where the boundary case is
  ambiguous, but you still want downstream logical operations to work.
- A boolean reduction of a column where you want to preserve the
  distinction between "genuinely true" and "true but with caveats."
- Any case where you would otherwise reach for a 4-state truth value
  (true, false, neither, both) but want the extra states to behave
  numerically like ordinary booleans.

`polean` introduces a non-propagating extension of the logical type that
preserves numerical equivalence with base R while adding two new
literals.

## The five literals

Every polean is one of five values:

| Literal    | Orientation | Strictness |
|------------|-------------|------------|
| `TRUE`     | true        | strict     |
| `FALSE`    | false       | strict     |
| `ISH(1)`   | true        | ish        |
| `ISH(0)`   | false       | ish        |
| `NA`       | NA          | NA         |

`TRUE` and `ISH(1)` share the same orientation (true) but differ in
strictness. Likewise for `FALSE` and `ISH(0)`. Internally, every polean
is stored as a single complex number: the real part is the orientation
(0, 1, or NA) and the imaginary part is the strictness flag (0 for
strict, 1 for ish).

## Two equality regimes

```{r}
polean(TRUE) == polean(TRUE)   # TRUE
polean(1) == polean(TRUE)      # FALSE: differ in strictness
polean(1) == 1                 # TRUE:  numerical equivalence
polean(TRUE) == 1              # TRUE:  numerical equivalence
```

Boolean equality (polean vs polean, or polean vs base `logical`)
requires both orientation *and* strictness to agree. Numerical equality
(polean vs base `numeric`) requires only orientation. The point is to
preserve the invariant that `polean(x) == x` whenever `x` is a numeric
or logical scalar, while still letting users distinguish strict from
ish at the polean level.

## Three ways to construct a polean

```{r}
polean(TRUE)         # TRUE        (logical input -> strict)
polean(1)            # ISH(1)      (numeric input -> ish)
polean(1+0i)         # TRUE        (complex with Im=0 -> strict)
polean(1+1i)         # ISH(1)      (complex with Im!=0 -> ish)

as.polean("TRUE")    # TRUE
as.polean("1.5")     # ISH(1)
as.polean(list(TRUE, 1, NA))

ISH(1)               # ISH(1)
ISH(TRUE)            # TRUE        (ISH delegates by R type)
```

`polean()` and `ISH()` are both wrappers around `as.polean()`. The
difference is that `ISH()` enforces a length-1 input, so it can be used
as a literal-style constructor; `polean()` accepts any length.

### Why is `polean(1)` ish, not strict?

A numeric `1` in user code is rarely meant as a boolean literal, but it
is often *used* in boolean-adjacent contexts (e.g. `sum(x == y)`,
flags from a `0/1` indicator vector). Treating it as ish marks the
fact that the user passed a number, not a logical, while keeping the
numerical comparison `polean(1) == 1` true. If you want a strict
`TRUE` from a numeric, use `polean(TRUE)`, `polean(1+0i)`, or
`as.polean("TRUE")`.

## Combination

```{r}
c(polean(TRUE), polean(1), NA)
c(polean(TRUE), 1, 1.5, FALSE)
```

When concatenating, polean inputs are preserved as-is and non-polean
inputs are coerced individually via `as.polean()`. Each element keeps
whatever orientation/strictness its R type implies.

## Display

```{r}
polean(c(TRUE, FALSE, NA))
polean(c(1, 0, NA))
c(polean(TRUE), ISH(1), NA, FALSE)
```

## Status

This package is in early development. The current release covers
construction, coercion, equality, and combination. Logical operators
(`&`, `|`, `!`) and arithmetic interactions are not yet defined; their
semantics under the strict/ish split are still being worked out.